Parallel Information Set Generation for Kriegspiel

### 2012-07-16 sbac/paper.tex
Changed a plot, muliple changes

\title{Parallel Information Set Generation for Kriegspiel 
       }

@@

In this work, we show that ISG for Kriegspiel can be efficiently parallelized.
The primary contributions of this work are the following:
\begin{itemize}
\item
 We describe an
approach to efficient parallelization of an application which is not typical of
high performance computing domain.
\item
 We present scalability results for our
parallel implementation of an information set generation algorithm for the game
of Kriegspiel using problem instances from real games. We achieve results which are unprecendented in game tree search application. 
\item
We demonstrate
the effects of load balancing strategy and grainsize on the parallel performance of our implementation. 
\end{itemize}

@@

\section{Information Set Generation}
\label{info}
%\subsection{Information Set Generation}
The concept of information sets is perhaps best explained through an example. 
Figures ~\ref{listing} show a full game listing and the associated pictorial progression of an actual
Kriegspiel game~\cite{li94chess}.  We have adopted the notation used by Wolfe for Berkeley Kriegspiel~\cite{wolfe07exploiting}.  Each numbered entry shows one complete turn for both White and Black.  The actual move is shown
first followed by a list of referee announcements.  Captures announcements are prefixed with an X and give the location
of the captured piece.  Similarly, pawn try announcements are prefixed with T.  Attempted moves that were ultimately
declared illegal are shown in a list following the `:'.

Figure~\ref{fig:figure1} shows the true state of the game after nine moves by
each player.  Figure~\ref{fig:figure2} shows White's perspective of the game at
that coin. At move 9, Black attempts to block a
potential threat along the diagonal by moving a pawn to g6, then attempts to
escape or capture a threat at f7.  When both of these attempts fail, Black can
infer that his king is being threatened by a protected bishop or queen at f7.
The attempt to move to e7 is Black's way of finding out whether the threatening
piece is a queen or bishop.  When this move fails, black knows it is White's
queen and ultimately retreats to d8.  Note that the failed moves by White on
turns 3 and 4 are a result of being blocked by the pawn at e2.

@@

\textbf{The Information Set Generation}
(ISG) operation can be formulated as a combinatorial search problem.  The input to the algorithm is a list
of $t$ observations for one player $k$ (denoted $o_{1:t}^{(k)}$ in a Kriegspiel game.

The nodes in the search tree at depth $n$ correspond possible to possible move 
sequences of $n$ plies from the start state.  (A sequence of $t$ actions from
the start state is denoted $a_{1:t}$.) There is a search space variable for
each ply in the game, and the values that a variable can take on are all
possible legal moves.  Information set generation, then, is an all-solutions
search through the space.  Whenever a node in the search tree is found to be
inconsistent with the actual observation given in the input, the subtree rooted
at that node is pruned (i.e., search is discontinued along that path).  There
are three types of observations that must match in order for a state $s$ to be
included in the information set $I$ at depth $t$: check status, pawn tries, and
illegal moves available/attempted.

\begin{figure}[htpb]
\small
%\begin{algorithm}
\begin{boxedminipage}{\columnwidth}
\begin{algorithmic}[1]
\Function{DFS-Infoset}{$a_{1:t}$, $o_{1:T}^{(k)}$, $I$, $s$, $t$}
  \If {$t > T$}
    \State {$I \leftarrow I \cup a_{1:t}$} %\Comment{$a_{1:t}$ leads to $o_{1:T}$ for $P_k$; add it to the information set.}
  \Else {
    \For{\textbf{each} legal action $a$ at $s$}
      \State {$s'$ = {\sc GetNextState}($s$,$a$)}\label{line:update}
      \If {$s'.o_{t+1}^{(k)}=o_{t+1}^{(k)}$}
        \State {{\sc DFS-Infoset}($a_{1:t}a$, $o_{1:T}^{(k)}$, $I$, $s'$, $t+1$)} 
          %\Comment{ Recall that $N(a_{1:t}a) = T(T(a_{1:t}),a)$}
      \EndIf
    \EndFor
  }\EndIf
\EndFunction
\end{algorithmic}
\end{boxedminipage}
\caption{Algorithm}
\label{codelisting1}
\end{figure}

@@

In contrast to other popular parallel programming frameworks such as {\sc MPI}
and {\sc OpenMP}, the {\sc Charm++} language is based on object-oriented
programming principles.  The programmer is responsible to decompose a problem
into a collection of objects that represent elements of work to be done.
Typically the number of these work objects, called {\em chares}, far exceeds
the number of processing elements (cores) available.  Chares are defined
so as to be able to be migrated between processors {\em during execution}, in
order to improve the overall load balancing.

Chares communicate with each other via
asynchronous message passing.  They can invoke {\em entry methods} on other
objects, which are executed according to a schedule set by the {\sc Charm++}
runtime system.  The runtime is responsible to interleave computation and
communication tasks in order to maximize efficiency.  To facilitate this, each
processor maintains an {\em incoming queue} of messages targeted to chares that
it currently holds.  A corresponding {\em outgoing queue} collects messages
generated by objects on the processor that need to be sent to chares assigned
to other processors.

@@

%\subsection{The {\sc Charm++} Parallel State Space Search Engine}
Sun \etal have developed the Parallel State Space Search Engine ({\sc ParSSSE})
within the {\sc Charm++} framework, which exposes an API that is useful for a
variety of graph-based and tree-based search problems~\cite{sun11adaptive}. In
{\sc ParSSSE} the fundamental unit of work is the processing of a node.
Generally, this means identifying whether the node corresponds to a solution
and/or generating descendant nodes and spawning work tasks related to them. A
task is {\em terminal} if it spawns no children.

A {\em grainsize control} parameter specifies a bound on the size of a piece of
work that can be passed along to the runtime system for scheduling.  Once a
task has been subdivided into an appropriate number of subtasks, those tasks
run to completion on a single processor without preemption and without generating
any additional parallel tasks.  If the grainsize is too large, then the number
of pieces of work to be done may be lower than the number of processors
available, forcing some processors to idle.  If the grainsize is too small,
then the runtime system is flooded with a large number of trivial tasks, and
the system gets bogged down by the overhead of creating and scheduling tasks
instead of actually doing them.  {\sc ParSSSE} supports adaptive grainsize
management, so that the threshold for sequential processing can be changed
during the execution of the program, depending on the properties of the problem
and/or the realtime performance analysis of the runtime system.

{\sc ParSSSE} currently supports two load balancing strategies.  The
first is the randomized strategy.  In this approach, each newly created object
is assigned to a random processor.  Generally, if a large enough number of 
chares is created, the system will achieve an approximately uniform
distribution of work across processors.  The approach is also appealing because
of its simplicity. However, randomization can sometimes lead to inefficiencies
from parallel overhead.  For a program running on $p$ processors, the
probability that a chare is assigned to a processor other than the
one that created it is $1-1/p$.  As the number of processors gets large, this
probability approaches 1, which means that the parallel overhead related to the
creation and scheduling of a new task is incurred for almost every piece of new
work.

@@

@@

\subsection{Analysis using Random Instance} 
The performance results reported in this section were obtained by running our
program on Blueprint with up to 1,024 processors. This system was composed of
120 POWER5+ nodes, each having 16 cores and 64GB of memory. The system was
decommissioned in June 2011 and is no longer in operation.

Table~\ref{speedups} shows the performance results for our algorithm on a
randomly selected problem instance in which the size of the information set was
about 4.5 million.  This instance is not necessarily representative of all
games or states of the game, but it was not cherry-picked in any way. The table
shows that the problem is indeed highly parallizable. The speedups are nearly
linear for up to 32 processors, and the speedups continue to increase even up
to 1,024 processors. In this particular instance, the sequential algorithm
takes almost 25 minutes to run.  But when parallelized on 1,024
processors---perfectly reasonable resource usage on modern hardware---the
search takes less than three seconds.  This is a speedup of 576 times! In a
tournament where each player is allotted an average of five minutes per move,
an ISG operation that requires 25 minutes or more would be prohibitively
expensive.  On the other hand, if the operation can be done in under three
seconds, ISG can be used as a subroutine in the decision-making process
(see~\cite{parker05game}).

@@

\caption{Speedups for a randomly selected problem instance on 1--1,024
processors.  The input observations were from the 10th ply of the game tree and
the size of the information set in this instance is 4,495,121.}
\centering
\begin{tabular}{rrr}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}

@@

@@

In our next round of experiments, we set out to compare load-balancing
strategies for the problem.  We used the random, work stealing, and
neighborhood balancing strategies that are available in the general purpose
search engine.  These results are shown in Figure~\ref{3schemes}.  Our results
show that there was actually little variation in performance across the
different strategies.  Across all of our experiments, the work stealing
approach appeared to be consistently---if only marginally---better than the
others.  Our hypothesis here is that the nodes are rarely starved for work.  In
other words, perhaps the very nature of the problem means that the load is
naturally well-balanced and that the load-balancer rarely needs to do a
significant amount of redistribution.  This is probably also related to the
fact that we are doing an all-solutions search. 
\begin{figure}[ht]
\centering
\includegraphics[width=0.9\columnwidth]{plots/3schemes.pdf}

@@

\subsubsection{Effect of Grainsize}

\begin{figure*}[!ht]
\centering
\subfigure[Depth 8]{
\includegraphics[width=2.1in]{plots/dep8.pdf}
\label{dep8}
} 
\subfigure[Depth 10]{
\includegraphics[width=2.1in]{plots/dep10.pdf}
\label{dep10}
} 
\subfigure[Depth 12]{
\includegraphics[width=2.1in]{plots/dep12.pdf}
\label{dep12}
} 
\caption{Grainsize Control for different problem sizes}
%\vspace{-0.1in}
\label{fig:dep}
\end{figure*}
%
\begin{figure}[ht]
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depths.pdf}
\end{minipage}
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depthsnolog.pdf}
\end{minipage}
\caption{(Top) Log-log plot of running time for three different sizes of problem as the number of cores used increases.
(Bottom) Same data with a linear y-axis for the running time, to help see the isoefficiency.}
\label{3depths}
\end{figure}

Next, we evaluate the effect of grainsize on execution time.  The results are
shown in Tables~\ref{dep8}--~\ref{dep12}.  It is clear that there is a ``sweet
spot'' for each problem size.  (At least it is clear for this particular
instance.  It is possible that there would be different behavior on different
problem instances.)  When the grainsize is too small, the overall computation
is hampered by excessive parallel overhead (communication, queueing of nodes to
expand, load balancing operations).  On the other hand, when the grainsize is
too large, it becomes too difficult to ensure that some processors are not
idling with no work to do while a small number of processors are busy doing
large chunks of work.

In our experiments, the optimal grainsize is clearly related to the size of the
problem, the depth of the search tree.  The optimal sequential threshold seems
to be $2-4$ less than the problem depth.  Since the depth of all solutions is
the same for each problem instance, this means that the number of levels that
should be searched sequentially is fairly consistent across problem sizes.  We
noted that the number of solutions found---in other words, the size of the
information sets---is typically fairly large with respect to the total number
of nodes expanded.  (We estimate that it was often $30-40\%$).  This high
density of solutions limits the variability that can be present in the amount
of work to be done at the nodes representing the last few levels of the tree.
This means that the best grainsize is also likely to be fairly predictable,
which is encouraging.

@@

We next study the variation in the efficiency of the parallelization of the
algorithm as size of the problem increases.  To test this, we varied the number
of plies in the test instance from 8 to 12. We used the work stealing
load-balancing strategy.

The results are shown in Figure~\ref{3depths}.  In the easiest case (depth 8),
there is not enough additional parallel work to justify the jump from 256 to
1,024 processors, and performance actually degrades.  Aside from this anomaly,
however, we see consistent speedups as we increase the number of processors.
The plot on the right helps to illustrate the good isoefficiency of the
problem.  As the size of the problem increases, the amount of parallelizable
work available also increases, and therefore the steepness of the speed curve
is higher for larger problem sizes, as the number of processors increases.

\subsection{Real Game Instance}
%\begin{figure}[ht]
% \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem4.pdf} %
%
%\caption{Kriegspiel position from Virgil-David 1969 with information set size
%764,209.}
%
%\label{prob4} 
%\vspace{-0.2in}
%\end{figure}
%
%\begin{figure}[ht]
% \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem5.pdf} %
%
%\caption{Kriegspiel position from Game 6 of Kbot-Darkboard at the 2006 Computer
%Chess championship.  Information set size is 5,099,257.}
%
%\label{prob5} 
%\vspace{-0.2in}
%\end{figure}

\begin{figure}[ht]
 \centering \includegraphics[scale=1]{plots/new.pdf} %
\caption{Kriegspiel position from 2 games, position 10}
\label{prob4} 
\end{figure}

@@

\centering
\begin{tabular}{rrrrr}
PE/G & 2 & 4 & 6 & 8 \\
1 & {\bf 177$|$1} & 178$|$0.99 & 1844$|$0.96 & 493$|$0.36 \\
4 & 71$|$2.48 & 53$|$3.34 & {\bf 48$|$3.63} & 75$|$2.35 \\
16 & 34$|$5.19 & 19 $|$9.27& {\bf 12$|$13.8} & 16$|$10.7 \\
64 & 24 $|$7.25 & 7 $|$23.6& {\bf 3.7$|$47.39} & 4.1$|$43.1 \\
256 & 24 $|$ 7.1& 4.4 $|$40.1 & 1.2 $|$145& {\bf 1.0$|$172} \\
1024 & 26 $|$6.61 & 3.8 $|$46.5& 0.64 $|$276& {\bf 0.40$|$443}
\end{tabular}

@@

\begin{table}[th]

\caption{Running times, in seconds and speedups (after $|$), to find full
information set of size 5,099,257 at position 10 of Game 6 Kbot-Darkboard 2006.
Optimal grainsize for each number of cores is shown in bold.}
\centering
\begin{tabular}{rrrrr}
PE/G & 2 & 4 & 6 & 8 \\
1 & 617$|$1 & {\bf 612$|$1} & 650$|$0.95 & 3312$|$0.19 \\
4 & 213$|$2.89 & 170$|$3.63 & {\bf 158$|$3.89} & 367$|$1.68 \\
16 & 83$|$7.41 & 49$|$12.4 & {\bf 40$|$15.1} & 64$|$9.57 \\
64 & 54$|$11.3 & 19$|$32.3 & {\bf 11$|$52.9} & 14.9$|$41.4 \\
256 & 54$|$11.3 & 9.8$|$62.9 & 3.5$|$172 & {\bf 3.5$|$174} \\
1024 & 58$|$10.6 & 5.5$|$111 & 1.3$|$464 & {\bf 1.2$|$536} \\
\end{tabular}

\label{tab:prob5}
\end{table}

The running times and speedups for two of the more challenging instances on
Ranger are shown in Tables~\ref{tab:prob4}--\ref{tab:prob5}.   These tables
show the results for runs on one to $1,024$ cores.  Each column corresponds to
a different grainsize parameter.  In this case, the grainsize is the number of
levels of the search tree that are searched in parallel.  Up to that threshold,
each node that is generated spawns a new piece of work, which the {\sc Charm++}
runtime schedules to execute on one of the processing elements according to its
load balancing strategy.  Above the threshold the entire subtree is searched to
completion on a single processor, with no additional parallel overhead.

Note that the optimal grainsize varies for the different numbers of cores and
the pattern is consistent across problem instances.  The optimal
grainsize increases as the number of cores increases.  This makes sense,
because as this parameter grows, the number of distinct pieces of work
increases.  When the number of chares per processor is high, much of
the overhead of parallelization is wasted.  But as the number of cores
increases, the work can be efficiently broken up into smaller pieces and doled
out to all the cores.  With more pieces of work, it becomes easier to balance
the load.

@@

Our results compare favorably to previous work in parallel game tree search,
shown in Table~\ref{bestspeedups}.  The previous best parallelization results
were for parallelizing alpha-beta pruning.  Ferguson and Korf achieved a
speedup of 12 on 32 cores in the game of Othello~\cite{ferguson88distributed}.
In the game of chess, Felton and Otto showed a speedup of 100 on 256
cores~\cite{felten88highly}.  This limited scalability is due to speculative
loss.  Parts of the game tree that are searched during the parallel execution
would have been pruned had the algorithm run sequentially.  This is a
fundamental limitation in alpha-beta search.

\begin{table}[ht]
\caption{Best Speedups reported for game tree applications.}
\centering
\begin{tabular}{lrrc}
{\bf Authors} & {\bf Game} & {\bf Year} & {\bf Best Speedup (cores)} \\
Ferguson \& Korf & Othello & 1988 & 12(32) \\
Felton \& Otto & Chess & 1988 & 100(256) \\
Richards \etal & Kriegspiel & 2012 & 178(256) \\
Richards \etal & Kriegspiel & 2012 & {\bf 536(1024)} \\
\end{tabular}
\label{bestspeedups}
\end{table}

@@

\bibliographystyle{IEEEtran}
\bibliography{paper}

>>>>

\title{Parallel Information Set Generation for Kriegspiel 
       }
\title{Parallelizing  Information Set Generation for Game Tree Search Applications}

@@

In this work, we show that ISG for Kriegspiel can be efficiently parallelized.
The primary contributions of this work are the following:
\begin{itemize}
\item
 We describe an
approach to efficient parallelization of an application which is not typical of
high performance computing and scientific domain, yet important because of its usefulness in the field of artificial intelligence.
\item
 We present scalability results for our
parallel implementation of an information set generation algorithm for the game
of Kriegspiel using problem instances from real games. We achieve results which are unprecedented in game tree search application. 
\item
We demonstrate
the effects of load balancing strategy and grainsize on the parallel performance of our implementation. 
\end{itemize}

@@

\section{Information Set Generation}
\label{info}
%\subsection{Information Set Generation}
The concept of information sets is perhaps best explained through an example. 
Figures~\ref{listing}, \ref{fig:figure1} and~\ref{fig:figure2} show a full game listing and the associated pictorial progression of an actual
Kriegspiel game~\cite{li94chess}.  We have adopted the notation used by Wolfe for Berkeley Kriegspiel~\cite{wolfe07exploiting}.  Each numbered entry shows one complete turn for both White and Black.  The actual move is shown
first followed by a list of referee announcements.  Capture announcements are prefixed with an X and give the location
of the captured piece.  Similarly, pawn try announcements are prefixed with T.  Attempted moves that were ultimately
declared illegal are shown in a list following the `:'.

Figure~\ref{fig:figure1} shows the true state of the game after nine moves by
each player.  Figure~\ref{fig:figure2} shows White's perspective of the game at
that point. At move 9, Black attempts to block a
potential threat along the diagonal by moving a pawn to g6, then attempts to
escape or capture a threat at f7.  When both of these attempts fail, Black can
infer that his king is being threatened by a protected bishop or queen at f7.
The attempt to move to e7 is Black's way of finding out whether the threatening
piece is a queen or bishop.  When this move fails, black knows it is White's
queen and ultimately retreats to d8.  Note that the failed moves by White on
turns 3 and 4 are a result of being blocked by the pawn at e2.

@@

\textbf{The Information Set Generation}
(ISG) operation can be formulated as a combinatorial search problem.  The input to the algorithm is a list
of $t$ observations for one player $k$ (denoted $o_{1:t}^{(k)}$ in a Kriegspiel game.    
\begin{figure}[htpb]
\small
%\begin{algorithm}
\begin{boxedminipage}{\columnwidth}
\begin{algorithmic}[1]
\Function{DFS-Infoset}{$a_{1:t}$, $o_{1:T}^{(k)}$, $I$, $s$, $t$}
  \If {$t > T$}
    \State {$I \leftarrow I \cup a_{1:t}$} %\Comment{$a_{1:t}$ leads to $o_{1:T}$ for $P_k$; add it to the information set.}
  \Else {
    \For{\textbf{each} legal action $a$ at $s$}
      \State {$s'$ = {\sc GetNextState}($s$,$a$)}\label{line:update}
      \If {$s'.o_{t+1}^{(k)}=o_{t+1}^{(k)}$}
        \State {{\sc DFS-Infoset}($a_{1:t}a$, $o_{1:T}^{(k)}$, $I$, $s'$, $t+1$)} 
          %\Comment{ Recall that $N(a_{1:t}a) = T(T(a_{1:t}),a)$}
      \EndIf
    \EndFor
  }\EndIf
\EndFunction
\end{algorithmic}
\end{boxedminipage}
\caption{Algorithm}
\label{codelisting1}
\end{figure}

The nodes in the search tree at depth $n$ correspond to possible move 
sequences of $n$ plies from the start state.  (A sequence of $t$ actions from
the start state is denoted $a_{1:t}$.) There is a search space variable for
each ply in the game, and the values that a variable can take on are all
possible legal moves.  Information set generation, then, is an all-solutions
search through the space.  Whenever a node in the search tree is found to be
inconsistent with the actual observation given in the input, the subtree rooted
at that node is pruned (i.e., search is discontinued along that path).  There
are three types of observations that must match in order for a state $s$ to be
included in the information set $I$ at depth $t$: check status, pawn tries, and
illegal moves available/attempted.

@@

In contrast to other popular parallel programming frameworks such as {\sc MPI}~\cite{MPI}
and {\sc OpenMP}~\cite{OpenMP}, {\sc Charm++} is based on object-oriented
programming principles.  The programmer is responsible to decompose a problem
into a collection of objects that represent elements of work to be done.
Typically the number of these work objects, called {\em chares}, far exceeds
the number of processing elements (cores) available.  Chares are defined
so as to be able to be migrated between processors {\em during execution}, in
order to improve the overall load balancing.

Chares communicate with each other via
asynchronous message passing.  They can invoke {\em entry methods} on other
objects, which are executed according to a schedule set by the {\sc Charm++}
runtime system.  The runtime is responsible for interleaving computation and
communication tasks in order to maximize efficiency.  To facilitate this, each
processor maintains an {\em incoming queue} of messages targeted to chares that
it currently holds.  A corresponding {\em outgoing queue} collects messages
generated by objects on the processor that need to be sent to chares assigned
to other processors.

@@

%\subsection{The {\sc Charm++} Parallel State Space Search Engine}
Sun \etal have developed the Parallel State Space Search Engine ({\sc ParSSSE})
within the {\sc Charm++} framework, which exposes an API that is useful for a
variety of graph-based and tree-based search problems~\cite{sun11adaptive}. In
{\sc ParSSSE} the fundamental unit of work is the processing of a tree node.
Generally, this means identifying whether the node corresponds to a solution
and/or generating descendant nodes and spawning work tasks related to them. A
task is {\em terminal} if it spawns no children.

A {\em grainsize control} parameter specifies a bound on the size of a piece of
work that can be passed along to the runtime system for scheduling (in other words, computational granularity of a task which will be performed sequentially).  Once a
task has been subdivided into an appropriate number of subtasks, those tasks
run to completion on a single processor without preemption and without generating
any additional parallel tasks.  If the grainsize is too large, then the number
of pieces of work to be done may be lower than the number of processors
available, forcing some processors to idle.  If the grainsize is too small,
then the runtime system is flooded with a large number of trivial tasks, and
application performance is degraded because of the overhead of creating and scheduling tasks
instead of actually doing them.  {\sc ParSSSE} supports adaptive grainsize
management, so that the threshold for sequential processing can be changed
during the execution of the program, depending on the properties of the problem
and/or the realtime performance analysis by the runtime system.

Charm++ (and hence {\sc ParSSSE}) provides multiple load balancing strategies and we used three of them in our application. We describe these strategies next. The
first is the randomized strategy.  In this approach, each newly created object
is assigned to a random processor.  Generally, if a large enough number of 
chares is created, the system will achieve an approximately uniform
distribution of work across processors.  The approach is also appealing because
of its simplicity. However, randomization can sometimes lead to inefficiencies
from parallel overhead.  For a program running on $p$ processors, the
probability that a chare is assigned to a processor other than the
one that created it is $1-1/p$.  As the number of processors gets large, this
probability approaches 1, which means that the parallel overhead related to the
creation and scheduling of a new task is incurred for almost every piece of new
work.

@@

Another strategy we evaluated is the neighbor based load balancing which imposes a virtual topology on the processors and load exchange occur between neighbors only. In this strategy, the load balancing is initiated by overloaded processors. When a processor becomes overloaded, it sends work to its neighbors.

@@

\subsection{Analysis using Random Instance}

@@

\caption{Speedups for a randomly selected problem instance on 1--1,024
processors.  The input observations were from the 10th ply of the game tree and
the size of the information set in this instance is 4,495,121.}
\centering
\begin{tabular}{ccc}
{\bf Procs}	&	{\bf Time (s)} 	&	{\bf Speedup}\\
1	&	1524	&	1\\
2	&	812	&	1.88\\
4	&	407	&	3.74\\
8	&	208	&	7.32\\
16	&	102	&	14.8\\
32	&	53.2	&	28.7\\
64	&	28.0	&	54.5\\
128	&	14.9	&	102\\
256	&	7.70	&	197\\
512	&	4.59	&	331\\
1024	&	2.64	&	576\\
\end{tabular}

@@

The performance results reported in this section were obtained by running our
program on Blueprint with up to 1,024 processors. This system was composed of
120 POWER5+ nodes, each having 16 cores and 64GB of memory. The system was
decommissioned in June 2011 and is no longer in operation.

Table~\ref{speedups} shows the performance results for our algorithm on a
randomly selected problem instance in which the size of the information set was
about 4.5 million.  This instance is not necessarily representative of all
games or states of the game, but it was not cherry-picked in any way. The table
shows that our implementation achieves good parallel scaling. The speedups are nearly
linear for up to 32 processors, and the speedups continue to increase even up
to 1,024 processors. In this particular instance, the sequential algorithm
takes almost 25 minutes to run.  However, when parallelized on 1,024
processors --- perfectly reasonable resource usage on modern hardware --- the
search takes less than three seconds.  This is a speedup of 576 times! In a
tournament where each player is allotted an average of five minutes per move,
an ISG operation that requires 25 minutes or more would be prohibitively
expensive.  On the other hand, if the operation can be done in under three
seconds, ISG can be used as a subroutine in the decision-making process
(see~\cite{parker05game}).

@@

In our next round of experiments, we evaluated the impact of load-balancing
strategies on our application's performance.  We used the random, work stealing, and
neighborhood balancing strategies that are available in the general purpose
search engine.  These results are shown in Figure~\ref{3schemes}.  Our results
show that there was actually little variation in performance across the
different strategies.  Across all of our experiments, the work stealing
approach appeared to be consistently --- if only marginally --- better than the
others.  Our hypothesis here is that the nodes are rarely starved for work.  In
other words, perhaps the very nature of the problem means that the load is
naturally well-balanced and that the load-balancer rarely needs to do a
significant amount of redistribution.  This is also related to the
fact that we are doing an all-solutions search. 
\begin{figure}[ht]
\centering
\includegraphics[width=0.9\columnwidth]{plots/3schemes.pdf}

@@

\subsubsection{Effect of Grainsize}
\label{grainsize}

\begin{figure*}[ht]
\centering
\subfigure[Depth 8]{
\includegraphics[width=2.1in]{plots/dep8.pdf}
\label{dep8}
} 
\subfigure[Depth 10]{
\includegraphics[width=2.1in]{plots/dep10.pdf}
\label{dep10}
} 
\subfigure[Depth 12]{
\includegraphics[width=2.1in]{plots/dep12.pdf}
\label{dep12}
} 
\caption{Grainsize Control for different problem sizes. Sequential threshold denoted the depth in the search tree after which the remainder of the task (game tree search) is performed sequentially. Grain size = max game tree depth - sequential threshold.}
%\vspace{-0.1in}
\label{fig:dep}
\end{figure*}
%

Next, we evaluate the effect of grainsize on execution time. 
Grainsize control is obtained in our implementation using a run-time argument called \emph{sequential threshold}, which specifies the depth in the search tree up to which parallel tasks would be created. With this terminology, grain size = max game tree depth - sequential threshold. Figures~\ref{dep8}--~\ref{dep12} illustrate the variation in application performance with sequential threshold for three different problem sizes.  It is clear that there is a ``sweet
spot'' for each problem size.  (At least it is clear for this particular
instance.  It is possible that there would be different behavior on different
problem instances.)  When the sequential threshold is large (grainsize is too small), the overall computation
is hampered by excessive parallel overhead (communication, queueing of nodes to
expand, load balancing operations).  On the other hand, when the sequential threshold is too small (grainsize is
too large), it becomes too difficult to ensure that some processors are not
idling with no work to do while a small number of processors are busy doing
large chunks of work.

In our experiments, the sequential threshold is clearly related to the size of the
problem --- the depth of the search tree.  The optimal sequential threshold is $2-4$ less than the problem depth.  Since the depth of all solutions is
the same for each problem instance, this means that the number of levels that
should be searched sequentially (that is, optimal grain size) is fairly consistent across problem sizes.  We
noted that the number of solutions found---in other words, the size of the
information sets---is typically fairly large with respect to the total number
of nodes expanded.  (We estimate that it was often $30-40\%$).  This high
density of solutions limits the variability that can be present in the amount
of work to be done at the nodes representing the last few levels of the tree.
This means that the best grainsize is also likely to be fairly predictable,
which is encouraging.

@@

We next study the variation in the efficiency of the parallelization of the
algorithm as size of the problem increases.  Here, we varied the number
of plies in the test instance from 8 to 12. We used the work stealing
load-balancing strategy.

The results are shown in Figure~\ref{3depths}.  In the easiest case (depth 8),
there is not enough additional parallel work to justify the jump from 256 to
1,024 processors, and performance actually degrades.  Aside from this anomaly,
however, we see consistent speedups as we increase the number of processors. 
The plot on the bottom helps to illustrate the effect of problem size on parallel efficiency (defined as Speedup/P) and helps understand the iso-efficiency of the problem. Iso-efficiency~\cite{b:grama-isoefficiency} means how should the problem size be increased with increasing number of processors to maintain same level of parallel efficiency. As the size of the problem increases, the amount of parallelizable work available also increases, and hence parallel efficiency is higher for larger problem sizes and constant number of processors (Figure~\ref{3depths} bottom). As the number of processors increases, parallel efficiency decreases for constant problem size. Figure~\ref{3depths} suggests that in parallel information set generation, to maintain constant parallel efficiency, problem size (search depth) should be roughly increased by two when the number of processor in increased by 4 times.
\begin{figure}[ht] 
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depths.pdf}
\end{minipage}
\centering
\begin{minipage}{0.7\linewidth}
\includegraphics[width=\columnwidth]{plots/3depthspe.pdf}
\end{minipage}
\caption{(Top) Execution time for three different sizes of problem as the number of cores used increases.
(Bottom) Parallel Efficiency (wrt to 16 cores) vs number of cores, to help see the effect of problem size and iso-efficiency}
\label{3depths}
\end{figure}
\subsection{Real Game Instance}
%\begin{figure}[ht]
% \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem4.pdf} %
%
%\caption{Kriegspiel position from Virgil-David 1969 with information set size
%764,209.}
%
%\label{prob4} 
%\vspace{-0.2in}
%\end{figure}
%
%\begin{figure}[ht]
% \centering \includegraphics[viewport= 280 200 300 550, scale=0.35]{images/KriegspielProblem5.pdf} %
%
%\caption{Kriegspiel position from Game 6 of Kbot-Darkboard at the 2006 Computer
%Chess championship.  Information set size is 5,099,257.}
%
%\label{prob5} 
%\vspace{-0.2in}
%\end{figure}

The running times and speedups for two of the more challenging instances are shown in Tables~\ref{tab:prob4}--\ref{tab:prob5}.   These tables
show the results for runs on one to $1,024$ cores.  Each column corresponds to
a different sequential threshold parameter - the number of
levels of the search tree that are searched in parallel.  Up to that threshold,
each node that is generated spawns a new piece of work, which the {\sc Charm++}
runtime schedules to execute on one of the processing elements according to its
load balancing strategy.  Beyond the threshold the entire subtree is searched to
completion on a single processor, with no additional parallel overhead.

These results confirm the observations made in Section~\ref{grainsize}. The optimal sequential threshold varies for the different numbers of cores and
the pattern is consistent across problem instances.  The optimal
sequential threshold increases as the number of cores increases.  This is expected,
because as this parameter grows, the number of distinct pieces of work
increases.  When the number of chares per processor is high, much of
the overhead of parallelization is wasted.  But as the number of cores
increases, the work can be efficiently broken up into smaller pieces and spread to all the cores.  With more pieces of work, it becomes easier to balance
the load.

@@

\centering
%\begin{tabular}{rrrrr}
\begin{tabular}{ccccc}
PE/G & 2 & 4 & 6 & 8 \\
1 & {\bf 177$|$1} & 178$|$0.99 & 1844$|$0.96 & 493$|$0.36 \\
4 & 71$|$2.48 & 53$|$3.34 & {\bf 48$|$3.63} & 75$|$2.35 \\
16 & 34$|$5.19 & 19 $|$9.27& {\bf 12$|$13.8} & 16$|$10.7 \\
64 & 24 $|$7.25 & 7 $|$23.6& {\bf 3.7$|$47.39} & 4.1$|$43.1 \\
256 & 24 $|$ 7.1& 4.4 $|$40.1 & 1.2 $|$145& {\bf 1.0$|$172} \\
1024 & 26 $|$6.61 & 3.8 $|$46.5& 0.64 $|$276& {\bf 0.40$|$443}
\end{tabular}

@@

\begin{table}[th]
\caption{Running times, in seconds and speedups (after $|$), to find full
information set of size 5,099,257 at position 10 of Game 6 Kbot-Darkboard 2006.
Optimal grainsize for each number of cores is shown in bold.}
\centering
%\begin{tabular}{rrrrr}
\begin{tabular}{ccccc}
PE/G & 2 & 4 & 6 & 8 \\
1 & 617$|$1 & {\bf 612$|$1} & 650$|$0.95 & 3312$|$0.19 \\
4 & 213$|$2.89 & 170$|$3.63 & {\bf 158$|$3.89} & 367$|$1.68 \\
16 & 83$|$7.41 & 49$|$12.4 & {\bf 40$|$15.1} & 64$|$9.57 \\
64 & 54$|$11.3 & 19$|$32.3 & {\bf 11$|$52.9} & 14.9$|$41.4 \\
256 & 54$|$11.3 & 9.8$|$62.9 & 3.5$|$172 & {\bf 3.5$|$174} \\
1024 & 58$|$10.6 & 5.5$|$111 & 1.3$|$464 & {\bf 1.2$|$536} \\
\end{tabular}

\label{tab:prob5}
\end{table}
\begin{figure}[ht]
 \centering \includegraphics[scale=1]{plots/new.pdf} %
\caption{Performance results for Information Set Generation for 2 games of Kriegspiel}
\label{prob4} 
\end{figure}

@@

\section{Comparison with previous results, Discussion}
Our results compare favorably to previous work in parallel game tree search,
shown in Table~\ref{bestspeedups}.  The previous best parallelization results
were for parallelizing alpha-beta pruning.  Ferguson and Korf achieved a
speedup of 12 on 32 cores in the game of Othello~\cite{ferguson88distributed}.
In the game of chess, Felton and Otto showed a speedup of 100 on 256
cores~\cite{felten88highly}.  This limited scalability is due to speculative
loss.  Parts of the game tree that are searched during the parallel execution
would have been pruned had the algorithm run sequentially.  This is a
fundamental limitation in alpha-beta search.

@@

\bibliographystyle{IEEEtran}
\bibliography{paper,cited}
